---
title: "Figure 2"
output: html_document
date: "2024-12-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Adapt from slide showing affine transformations, but include an example of canonical orientation
2. Example of such an alignment, can use Xenium to Visium for the lung cancer dataset I’m currently working on, spoiler to the next paper
3. Move the myofiber and Visium spot plot from the old Voyager paper
4. Spatial binning schematic
5. Contiguity splitting example
6. Extract raster values with vector; real life usecase in QC to find Visium spots with less tissue in them


```{r}
library(SpatialFeatureExperiment)
library(SFEData)
library(Voyager)
library(tidyverse)
library(scuttle)
library(sf)
library(scater)
library(DelayedMatrixStats)
library(terra)
library(EBImage)
library(BiocParallel)
theme_set(theme_bw())
```

# Affine
```{r}
if (!dir.exists(file.path("outs", "spatial"))) {
    download.file("https://cf.10xgenomics.com/samples/spatial-exp/2.0.0/Visium_Mouse_Olfactory_Bulb/Visium_Mouse_Olfactory_Bulb_spatial.tar.gz",
                  destfile = file.path("outs", "spatial.tar.gz"))
    untar(file.path("outs", "spatial.tar.gz"), exdir = "outs")
}
```

```{r}
sfe_ob <- read10xVisiumSFE(".")
```

```{r}
(p1 <- plotGeometry(sfe_ob, colGeometryName = "spotPoly", image_id = "lowres", fill = FALSE))
```

```{r}
ggsave("fig2/p1a.jpg", p1, width = 2, height = 3)
```

```{r}
sfe_ob2 <- SpatialFeatureExperiment::rotate(sfe_ob, 90, sample_id = "all")
```

```{r}
(p1b <- plotGeometry(sfe_ob2, colGeometryName = "spotPoly", image_id = "lowres", fill = FALSE))
```

```{r}
ggsave("fig2/p1b.jpg", p1b, width = 3, height = 2)
```

```{r}
bbox <- c(xmin=3400, xmax=4400, ymin=3000, ymax=4000)
bbox_sf <- st_bbox(bbox) |> st_as_sfc()
```

```{r}
(p2 <- plotGeometry(sfe_ob2, colGeometryName = "spotPoly", image_id = "lowres", fill = FALSE) +
     geom_sf(data = bbox_sf, color = "red", fill = NA, linewidth = 1))
```

```{r}
ggsave("fig2/p2a.jpg", p2, width = 3, height = 2)
```

```{r}
(p2b <- plotGeometry(sfe_ob2, colGeometryName = "spotPoly", image_id = "hires", fill = FALSE,
                     bbox = bbox))
```

```{r}
ggsave("fig2/p2b.jpg", p2b, width = 2, height = 2)
```


# Affine alignment

```{r}
sfe1 <- read10xVisiumSFE(samples = "TSU-21", dirs = "~/hd/data/Haga/SpatialOmics/Visium/TSU-21", 
                         data = "filtered", zero.policy = TRUE, unit = "micron")
```

```{r}
sfe2 <- readXenium("~/hd/data/Haga/SpatialOmics/Xenium/TSU-21/")
```

```{r}
plotImage(sfe1, image_id = "lowres")
```

```{r}
plotImage(sfe2, image_id = "morphology_focus")
```

```{r}
img1 <- toExtImage(getImg(sfe1, image_id = "hires"))
img2 <- toExtImage(getImg(sfe2, image_id = "morphology_focus"), resolution = 6)
```

```{r}
dim(img1)
```

```{r}
dim(img2)
```

```{r}
bar <- clahe(normalize(img2), nx = 4, ny = 6)
```

```{r}
display(bar, method = "raster")
```

```{r}
img1 <- normalize(img1)
colorMode(img1) <- Grayscale
img2 <- bar
```

```{r}
mask1 <- img1[,,2] < otsu(img1[,,2])
mask2 <- img2 > otsu(img2)
```

```{r}
mask1 <- fillHull(mask1)
```

```{r}
display(mask1)
```

```{r}
mask2 <- closing(mask2, makeBrush(7))
mask2 <- fillHull(mask2)
```

```{r}
display(mask2)
```

```{r}
# It's like every project has its own version of this
get_tissue_boundary <- function(mask, n_pieces = 1, simplify = TRUE, ...) {
    mask <- toSpatRasterImage(mask, save_geotiff = FALSE)
    names(mask) <- "value"
    polys <- as.polygons(mask)
    polys <- polys[polys$value==1,]
    polys <- st_as_sf(polys) |> st_geometry()
    polys <- st_cast(polys, "POLYGON")
    area <- st_area(polys)
    if (n_pieces < length(area)) 
        polys <- polys[order(area, decreasing = TRUE)[seq_len(n_pieces)]]
    if (simplify) {
        #polys <- rmapshaper::ms_simplify(polys, ...)
        polys <- st_simplify(polys, ...)
    }
    polys
}
```

```{r}
tb1 <- get_tissue_boundary(mask1, dTolerance=50)
tb2 <- get_tissue_boundary(mask2, dTolerance=50)
```

```{r}
ggplot() + geom_sf(data = tb1, color = "red", fill = NA) + 
    geom_sf(data = tb2, color = "blue", fill = NA)
```


```{r}
.npts <- function(x) nrow(st_coordinates(x))
```

```{r}
.npts(tb2)
```

```{r}
library(LOMAR)
```

```{r}
verts1 <- st_coordinates(tb1)[,c("X", "Y")]
verts2 <- st_coordinates(tb2)[,c("X", "Y")]
```

```{r}
trans <- cpd(verts2, verts1, scale = FALSE)
```

```{r}
v_trans2 <- as.data.frame(trans$Y)
v_trans2$type <- "transformed"
v_ref2 <- as.data.frame(verts2)
v_ref2$type <- "reference"
v <- rbind(v_ref2, v_trans2)
names(v) <- c("X", "Y", "type")
```

```{r}
ggplot(v, aes(X, Y, color = type)) +
    geom_point(size = 0.5) +
    geom_path() +
    coord_equal()
```

Not bad. Using sf's st_simplify worked better than using rmapshaper, probably because it preserves more points at the curvier places.
```{r}
trans[c("R", "t", "s")]
```

```{r}
get_transform <- function(moving, fixed, R, t) {
    mu_y <- colMeans(moving)
    mu_x <- colMeans(fixed)
    t_use <- -R %*% mu_y + t + mu_x
    list(M = R, v = as.vector(t_use))
}
```

```{r}
ts <- get_transform(verts1, verts2, trans$R, trans$t)
```

```{r}
foo <- t(ts$M %*% t(verts1) + ts$v)
```

```{r}
as.data.frame(foo) |> 
    ggplot(aes(V1, V2)) + geom_point() + coord_equal() +
    geom_path(data = v, aes(X, Y, color = type))
```

```{r}
foo <- df2sf(foo, spatialCoordsNames = c("V1", "V2"))
```

```{r}
sfe1_t <- SpatialFeatureExperiment::affine(sfe1, M = ts$M, v = ts$v)
```

```{r}
library(alabaster.sfe)
```

```{r}
saveObject(sfe1_t, "haga21_visium")
```

```{r}
(p7a <- plotImage(sfe1_t, image_id = "lowres", 
                  bbox = as.vector(st_bbox(foo)) |> 
                      setNames(c("xmin", "ymin", "xmax", "ymax"))))
```

```{r}
ggsave("fig2/p7a.jpg", p7a, width = 1.5, height = 3)
```

OK, so it works. Then what? How should the figure look? Overlaying the Visium spots on Xenium cells doesn't really make a good figure. I can just show the transformed image

```{r}
plotGeometry(sfe1, colGeometryName = "spotPoly", fill = FALSE,
             image_id = "lowres")
```

```{r}
(p7c <- plotGeometry(sfe1_t, colGeometryName = "spotPoly", fill = FALSE,
             image_id = "hires") +
    geom_sf(data = tb2, fill = NA, color = "blue"))
```

```{r}
writeImage(bar, "fig2/p7b.jpg", quality = 95)
```

```{r}
ggsave("fig2/p7c.jpg", p7c, width = 2, height = 4)
```

```{r}
dim(sfe1)
```

```{r}
cg <- spotPoly(sfe1_t)
```

```{r}
sfe3 <- aggregateTxTech("~/hd/data/Haga/SpatialOmics/Xenium/TSU-21/",
                        by = spotPoly(sfe1_t), tech = "Xenium",
                        BPPARAM = SerialParam(progressbar = TRUE))
```

```{r}
sfe1_t <- sfe1_t[rowData(sfe1_t)$symbol %in% rownames(sfe3),]
```

```{r}
sfe3 <- addPerCellQC(sfe3, subsets = list(visium = rownames(sfe3) %in% rowData(sfe1_t)$symbol))
sfe1_t <- addPerCellQC(sfe1_t)
```

```{r}
plotSpatialFeature(sfe1_t, "sum", colGeometryName = "spotPoly")
```

```{r}
plotSpatialFeature(sfe3, "subsets_visium_sum", colGeometryName = "bins")
```

Weird that there's such difference between Visium and Xenium. I wonder if it happens to other datasets as well. For the intersecting genes, even the general patterns are different and I can see a lot more structure on Xenium.

```{r}
saveObject(sfe3, "haga21_xenium_pseudo_visium")
```


# Myofiber

```{r}
sfe1 <- McKellarMuscleData("full")
```

```{r}
sfe_tissue <- sfe1[,colData(sfe1)$in_tissue]
sfe_tissue <- sfe_tissue[rowSums(counts(sfe_tissue)) > 0,]
```

```{r}
(p3a <- plotSpatialFeature(sfe_tissue, features = "nCounts", 
                           colGeometryName = "spotPoly",
                           annotGeometryName = "myofiber_simplified", 
                           aes_use = "color", linewidth = 0.5, fill = NA,
                           annot_aes = list(fill = "area")))
```

```{r}
ggsave("fig2/p3a.png", p3a, width = 3.5, height = 4)
```

```{r}
colData(sfe_tissue)$mean_myofiber_area <- 
  annotSummary(sfe_tissue, "spotPoly", "myofiber_simplified", 
               annotColNames = "area")[,1] # it always returns a data frame
```

```{r}
(p3b <- plotColData(sfe_tissue, x = "nCounts", y = "nGenes", colour_by = "mean_myofiber_area"))
```

```{r}
ggsave("fig2/p3b.png", p3b, width = 5, height = 3.5)
```

# Spatial binning schematic

```{r}
fp <- XeniumOutput("v2")
fp <- "xenium2"
sfe <- readXenium(fp, add_molecules = TRUE)
```

```{r}
rv <- rowVars(counts(sfe))
rs <- rowSums(counts(sfe))
gene_use <- rownames(sfe)[which.max(rs)]
```

```{r}
grid <- st_make_grid(centroids(sfe), cellsize = 55, square = FALSE)
```

```{r, fig.width=3.5, fig.height=3}
(p4a <- plotSpatialFeature(sfe, gene_use, colGeometryName = "centroids", size = 0.3, exprs_values = "counts") +
    geom_sf(data = grid, color = "gray50", fill = NA))
```

```{r}
ggsave("fig2/p4a.png", p4a, width = 3.5, height = 3)
```

```{r}
sfe_bc <- aggregate(sfe, by = grid, colGeometryName = "centroids")
```

```{r}
(p4b <- plotSpatialFeature(sfe_bc, gene_use, exprs_values = "counts"))
```

```{r}
ggsave("fig2/p4b.png", p4b, width = 3.5, height = 3)
```

```{r}
bbox2 <- c(xmin=700, xmax=850, ymin=-800, ymax=-650)
sfe_sub <- crop(sfe, bbox2)
```

```{r}
gene_use2 <- rownames(sfe_sub)[which.max(rowSums(counts(sfe_sub)))]
```

```{r}
grid2 <- st_make_grid(txSpots(sfe_sub), cellsize = 20, square = FALSE)
```

```{r}
(p4c <- plotGeometry(sfe_sub, colGeometryName = "cellSeg", rowGeometryName = "txSpots", gene = gene_use2,
             fill = FALSE) +
     geom_sf(data = grid2, fill = NA, color = "cornflowerblue"))
```

```{r}
ggsave("fig2/p4c.png", p4c, width = 3, height = 3)
```

```{r}
txSpots(sfe_sub) <- st_cast(txSpots(sfe_sub), "MULTIPOINT") |> st_zm()
```

```{r}
sfe_agg_cell <- aggregate(sfe_sub, cellSeg(sfe_sub), rowGeometryName = "txSpots")
```

```{r}
(p4d <- plotSpatialFeature(sfe_agg_cell, features = gene_use2, exprs_values = "counts",
                   rowGeometryName = "txSpots", rowGeometryFeatures = gene_use2))
```

```{r}
ggsave("fig2/p4d.png", p4d, width = 3.5, height = 3)
```

```{r}
sfe_agg_grid <- aggregate(sfe_sub, rowGeometryName = "txSpots", cellsize = 20, square = FALSE)
```

```{r}
(p4e <- plotSpatialFeature(sfe_agg_grid, features = gene_use2, exprs_values = "counts",
                   rowGeometryName = "txSpots", rowGeometryFeatures = gene_use2))
```

```{r}
ggsave("fig2/p4e.png", p4e, width = 3.5, height = 3)
```

# Contiguity splitting
```{r}
pieces <- readRDS(system.file("extdata/pieces.rds", package = "SpatialFeatureExperiment"))
pieces <- st_sf(geometry = pieces, sample_id = "sample01")
annotGeometry(sfe, "pieces") <- pieces
```

```{r}
(p5a <- plotGeometry(sfe, colGeometryName = "cellSeg", annotGeometryName = "pieces"))
```

```{r}
ggsave("fig2/p5a.png", p5a, width = 4, height = 3)
```

```{r}
sfes_pieces <- splitByCol(sfe, pieces)
```

```{r}
# Combine them back as different samples
sfes_pieces[[2]] <- changeSampleIDs(sfes_pieces[[2]], c(sample01 = "sample02"))
sfe2 <- do.call(cbind, sfes_pieces)
(p5b <- plotGeometry(sfe2, colGeometryName = "cellSeg"))
```

```{r}
ggsave("fig2/p5b.png", p5b, width = 6, height = 3)
```

# Raster to vector
```{r}
imgData(sfe_ob2)
```

```{r}
Img(sfe_ob2, image_id = "hires") <- toSpatRasterImage(getImg(sfe_ob2, image_id = "hires"), 
                                                      save_geotiff = FALSE)
```

```{r}
df <- extract(getImg(sfe_ob2, image_id = "hires"), vect(spotPoly(sfe_ob2)))
```

```{r}
df <- df |> 
    mutate(mean = rowMeans(select(df, starts_with("lyr")))) |> 
    group_by(ID) |> 
    summarize(across(everything(), mean))
```

```{r}
sfe_ob2$mean_value <- df$mean
```

```{r}
plotSpatialFeature(sfe_ob2, features = "mean_value")
```

```{r}
sfe_ob2 <- addPerCellQC(sfe_ob2)
```

```{r}
library(mclust)
```

```{r}
clusts <- Mclust(colData(sfe_ob2)[,c("mean_value", "sum")])
```

```{r}
summary(clusts)
```

```{r}
clusts$G
```

```{r}
sfe_ob2$clusts <- as.character(unname(clusts$classification))
```

```{r}
data("ditto_colors")
```

```{r}
(p6a <- plotColData(sfe_ob2, x = "mean_value", y = "sum", color_by = "clusts") +
    geom_density2d() +
    scale_color_manual(values = ditto_colors, name = "cluster"))
```

```{r}
ggsave("fig2/p6a.png", p6a, width = 5, height = 3.5)
```

```{r}
(p6b <- plotSpatialFeature(sfe_ob2, features = "clusts", image_id = "lowres") +
     scale_fill_manual(values = ditto_colors, name = "cluster"))
```

```{r}
ggsave("fig2/p6b.png", p6b, width = 5, height = 3)
```

